\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{hyperref}
\usepackage{authblk}

\setlength\parindent{0pt}
\setlength{\parskip}{8pt}

\title{Project Proposal: Benchmarking Monte Carlo Methods for Asian Options}

\author{Rohan Yelandur}
\author{Arnav Balaji} 
\author{Matthew Johnson}
\affil{Kernel Sanders (14)}

\date{\today}
\begin{document}

\maketitle

\section{Project Description}

This project implements and optimizes a Monte Carlo simulator for pricing Asian options in
the C programming language. We will begin with a correct but naive, single-threaded baseline
implementation. We then apply five high-performance computing optimizations:
compiler flag tuning, cache-friendly memory layout, SIMD vectorization, parallelism, and replacement of \texttt{rand()} with
thread-safe random number generation techniques. The goal is a high-throughput simulation engine where the impact
of each improvement is measured in isolation, with both performance and statistical
correctness analyzed at each step.

\section{Background and Relevance}

An option is a financial contract that gives the buyer the right to buy or sell an underlying
asset at a predetermined price. Asian options are a specific type of option whose final
payoff is based on the average price of the asset over the option's lifetime, rather than
just the ending price. ``Pricing'' an option means determining a fair value at which
buyers and sellers initiate a transaction, a concept essential to trillions of dollars of
daily market activity. Asian options are especially important in thinly-traded commodity
markets such as crude oil, where averaging over time prevents price manipulation near
maturity and reduces hedging costs relative to standard European options~\cite{Han2022ReviewOA}.

Standard options can be priced using the Black-Scholes closed-form formula, which models the
underlying asset as a geometric Brownian motion. However, no closed-form solution
exists for Asian options because the arithmetic average of lognormally distributed prices
does not itself follow a lognormal distribution~\cite{Han2022ReviewOA}. Closed-form approximations
using geometric averages exist but systematically underestimate the true price and lose
accuracy under high volatility~\cite{Han2022ReviewOA}.


A suitable numerical solution is the Monte Carlo method, which estimates the value of an
uncertain quantity by averaging the outcomes of many independent random
simulations~\cite{Zhang2009PricingAO}. When applied to Asian options, each simulation generates a
plausible price path for the underlying asset, then computes the payoff for that path. By the law of
large numbers, the average payoff across all paths converges to the true option price. Since
each path is independent, estimation error decreases at a rate of $1/\sqrt{n}$, where $n$
is the number of simulated paths~\cite{Zhang2009PricingAO}.

Achieving high simulation throughput is essential for financial institutions that must price
and re-price options in real time to capture market inefficiencies and manage risk. The
computational challenge arises because each of the millions of independent paths requires a
long sequence of random number generations and floating-point operations, making the
workload simultaneously compute-intensive, memory-intensive, and randomness-intensive. While
the structure of path simulation distributes naturally across cores,
a naive implementation using \texttt{rand()} isn't thread-safe. Concurrent calls introduce
correlations between paths that bias the final estimate~\cite{Bradley2011ParallelizationTF}. Further
performance optimizations are left on the table through poor memory layouts, unexploited SIMD hardware, and
untuned compiler flags. These challenges make monte carlo simulations a
natural fit for applying HPC improvements.

\section{Expected Results}

We will produce the following deliverables:

\begin{itemize}
    % \item \textbf{Formally specified problem.} Pricing Asian call options under the
    % risk-neutral geometric Brownian motion framework, where the underlying asset price
    % follows
    % \[
    %     dS = rS\,dt + \sigma S\,dW(t),
    % \]
    % with parameters $S_0$ (initial price), $K$ (strike), $r$ (risk-free rate), $\sigma$
    % (volatility), $T$ (maturity), $N$ (time steps per path), and $M$ (number of paths). The
    % Monte Carlo estimator and its $1/\sqrt{M}$ convergence guarantee will be stated
    % formally~\cite{kemna1990, Han2022ReviewOA}.

    \item \textbf{Implementation.} A C codebase containing a single-threaded baseline
    compiled at \texttt{-O0} and five incremental optimized versions, each differing by
    exactly one improvement to allow clean speedup attribution.

    \item \textbf{Correctness argument.} A proof that the parallelized estimator remains
    statistically unbiased and that the per-thread RNG streams are independent. All
    implementations will be verified to converge to the same reference price within a
    statistical tolerance.

    \item \textbf{Performance analysis.} A speedup table and roofline plot for all six
    versions. We will include thread scaling data from 1 to all available cores on a UTCS
    lab machine, along with GFLOPS/s estimates and memory bandwidth measurements.

    \item \textbf{Stretch goals.} A CUDA implementation that offloads path
    simulation to the GPU, exploiting the parallel outer loop~\cite{Bradley2011ParallelizationTF}.
    Also an exploration of Quasi-Monte Carlo methods and variance reduction techniques. ~\cite{Zhang2009PricingAO}. These address
    statistical efficiency rather than hardware throughput.
\end{itemize}

We define success as all implementations producing prices within a reasonable statistical
margin of the reference value. We expect the fully optimized version to achieve at least a
4$\times$ speedup over the \texttt{-O0} baseline, with each individual optimization
contributing a measurable improvement.

\section*{Team Plan}

All of us will contribute across every part of the project, but each will take
primary responsibility for one area.

\textbf{Rohan: } prove that each HPC optimization preserves mathematical correctness. This
includes showing that the parallelized Monte Carlo estimator is statistically unbiased,
that RNG is properly independent per thread. He will also help design the cache-friendly memory layout for the C implementation.

\textbf{Matthew: } write the C codebase including the \texttt{-O0} baseline, all five incrementally optimized versions, SIMD vectorization via AVX2 intrinsics, OpenMP parallelism, and the parallel RNG replacement.

\textbf{Arnav: } build the benchmarking infrastructure and testing scripts. He will run all implementations across varying thread counts and problem sizes, collect GFLOPS/s and memory bandwidth data, and write the performance analysis section of the final report.

\bibliographystyle{plain}
\bibliography{citations}

\end{document}